Predicting tree failure likelihood for utility risk mitigation via a novel convolutional neural network

### 2021-02-09 docs/manuscript/manuscript.tex
finished adding plot function to preprocess-images.py

Individual tree risk assessment can be costly because of the time it requires. In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{e.thomassmiley2017best}. Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency}. The utility of Level 1 assessments in these states suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.

@@

\section{Data and Methods}
\subsection{Image data description}
The training dataset consisted of 505 images, each having an original size of $4032\times3024$ pixels.   Images were captured over a single field season in Massachusetts, USA, between May and September 2020 to limit any potential influence of changes in tree appearance due to seasonal leaf senescence on image processing.  ESRI ArcMaps was used to randomly distribute sampling sites across the state.  Likelihood of failure assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.  This method is commonly used to assess trees in the United States.  A level 1 assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e. Level 2 or Level 3 \cite{e.thomassmiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{johnw.goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.  The four categories of likelihood of tree failure, which are always considered in a stated time frame, are defined as follows \cite{e.thomassmiley2017best}:
\begin{itemize}
\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
\item \textbf{Probable}: failure expected under normal weather conditions within a given time frame
\item \textbf{Imminent}: failure has started or is most likely to occur in the near future, even if there is no significant wind or increased load. This is a rare occurrence for a risk assessor to encounter, and may require immediate action to protect people from harm
\end{itemize}
	In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of trees in the \textit{imminent} category.

@@

>>>>

Individual tree risk assessment can be costly because of the time it requires. In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{smiley2017best}. Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency}. The utility of Level 1 assessments in these states suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.

@@

\section{Data and Methods}
\subsection{Image data description}
The training dataset consisted of 505 images, each having an original size of $4032\times3024$ pixels.   Images were captured over a single field season in Massachusetts, USA, between May and September 2020 to limit any potential influence of changes in tree appearance due to seasonal leaf senescence on image processing.  ESRI ArcMaps was used to randomly distribute sampling sites across the state.  Likelihood of failure assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{smiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{goodfellow2020best}.  This method is commonly used to assess trees in the United States.  A level 1 assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e. Level 2 or Level 3 \cite{smiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.  The four categories of likelihood of tree failure, which are always considered in a stated time frame, are defined as follows \cite{smiley2017best}:
\begin{itemize}
\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
\item \textbf{Probable}: failure expected under normal weather conditions within a given time frame
\item \textbf{Imminent}: failure has started or is most likely to occur in the near future, even if there is no significant wind or increased load. This is a rare occurrence for a risk assessor to encounter, and may require immediate action to protect people from harm
\end{itemize}
In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of trees in the \textit{imminent} category. Two randomly chosen examples are shown for each category in \autoref{fig:raw_images}.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.7\textwidth]{raw_input_images_2}
  \caption{Random examples of the raw training images (2 are shown per class). Images are antialiased in this figure for greater clarity.}
  \label{fig:raw_images}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{processed_input_images_Pr_Po_Im_64_px}
    \caption{Scenario \texttt{Pr\_Po\_Im}}
    \label{pr_po_im_64}
  \end{subfigure}%
  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{processed_input_images_Pr_Im_64_px}
    \caption{Scenario \texttt{Pr\_Po\_Im}}
    \label{pr_im_64}
  \end{subfigure}%

  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{processed_input_images_PrPo_Im_64_px}
    \caption{Scenario \texttt{Pr\_Po\_Im}}
    \label{prpo_im_64}
  \end{subfigure}%
  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{processed_input_images_Pr_PoIm_64_px}
    \caption{Scenario \texttt{Pr\_Po\_Im}}
    \label{pr_poim_64}
  \end{subfigure}%

  \caption{Random examples of the processed training images under each classification scenario; 64 pixels.}
  \label{fig:processed_images}
\end{figure}
